---
title: "Global Warming in Saxony?"
output:
    html_document:
        toc: true
        toc_float: true
        theme: cosmo
        highlight: tango
output_options: "self_contained"
---

To demonstrate the expressiveness and use of some bits and pieces of tidyverse, I'd like to reproduce the following graphic:
[this](https://twitter.com/MetOffice_Sci/status/1266361148527370242?s=20) posted of the UK MET service on May 29, 2020, on twitter. Out goal will be to try and reproduce it.


# First look into the dataset



```{r load_data}
library(dplyr, warn.conflicts=FALSE)
library(lubridate, warn.conflicts=FALSE)
library(readr)
library(tidyr)

rawdf = read.csv("saxony-monthly-temperature.csv")
glimpse(rawdf)
```
## create columns for year and month

```{r tidydate}
library(tidyr)

rawdf = rawdf %>% 
     mutate(date=ymd_hms(MESS_DATUM)) %>% 
     mutate(yr=as.integer(year(date)), mo=as.integer(month(date)))%>% 
     select(STATIONS_ID,MESS_DATUM,name,mo,yr,MO_TT)
glimpse(rawdf)
     
```

## Range of datasets

```{r maxmin2_recording}
message(paste("recording started",
           min(date(rawdf$MESS_DATUM)),
           "and ended",
           max(date(rawdf$MESS_DATUM))
           )
           )
```

## which stations recorded before 1960 or after 1990

```{r longstations}
library(tidyr)

refdf = rawdf %>% 
    filter(yr > 1959 & yr < 1991)

longstations = refdf %>%
     distinct(name)
     
message("found ",length(longstations$name)," / ", length(unique(rawdf$name)),
        " stations that recorded in of reference period")

df = rawdf %>% filter(name %in% longstations$name)
message("using ", length(unique(df$name))," stations (",length(unique(df$STATIONS_ID)),")")
     
```


## let's tidy up a bit

```{r tidyNAs}
library(knitr)

rawnas = df %>% filter(is.na(MO_TT))
     
summna = df %>% filter(is.na(MO_TT)) %>%
            group_by(STATIONS_ID) %>%
            summarize(cnt = n()) %>%
            mutate( maxposs = as.integer(12*length(unique(refdf$yr))),
            naratio = cnt/maxposs)
            
#kable(summna)

message(paste("stations with NAs:",length(unique(rawnas$STATIONS_ID)),"/",length(unique(df$STATIONS_ID))))
message(paste("stations with less than 25% NAs:",summna %>% filter(naratio < .25) %>% nrow()))

todiscard = summna %>% filter(naratio > .25)
df = df %>% filter( ! (STATIONS_ID %in% todiscard$STATIONS_ID) )

kable(df %>% filter(STATIONS_ID == 965, yr > 1978 & yr<1981))

oldcnt = df %>% filter( is.na(MO_TT) ) %>% nrow()
df = df %>% 
     group_by(STATIONS_ID,mo) %>%
     fill(MO_TT, .direction="downup") %>%
     ungroup()
     #filter(! is.na(MO_TT)) 
     
kable(df %>% filter(STATIONS_ID == 965, yr > 1978 & yr<1981))

newcnt = df %>% filter( is.na(MO_TT) ) %>% nrow()
message(paste("fixing NAs:",oldcnt,newcnt))

glimpse(df)
```


## calculate the average temperature within 1961-1990

We can calculate the reference median monthly temperature for each weather station.

```{r per_station_ref}
statdf = df %>% 
     filter(yr > 1960 & yr < 1991) %>%
     group_by(mo,STATIONS_ID) %>% 
     summarize(medTT = median(MO_TT), 
     name=unique(name))

message(paste("using",length(unique(statdf$STATIONS_ID)),"stations by id and",length(unique(statdf$name))), " by name")

if( T %in% is.na(statdf$medTT)) {
 message(paste("WARNING, reference temperatures contain NA values, original contained",df %>% filter(is.na(MO_TT)) %>% nrow()))
 cat(paste(is.na(statdf$medTT)))
} else {
kable(statdf %>% filter(STATIONS_ID == 840) )
}
```

# focussing on the annual deviation

First, we limit the dataset to those stations that yield a reference measurement.

```{r only_stations_from_ref}

df = df %>% filter( STATIONS_ID %in% statdf$STATIONS_ID | name %in% statdf$name)
message(paste("using",length(unique(df$STATIONS_ID)),"stations by id and",length(unique(df$name))), " by name")

if( T %in% is.na(df$MO_TT)) {
 message(paste("WARNING, reference temperatures contain NA values, original contained",df %>% filter(is.na(MO_TT)) %>% nrow()))
 cat(paste(is.na(df$MO_TT)))
}
```

Calculate the annual deviation from the reference temperature

```{r create_diff_to_ref}
olddf = df
small = statdf %>% select(STATIONS_ID, mo, medTT)
df = df %>% 
     inner_join(small, by=c("STATIONS_ID","mo")) %>% 
     mutate(MO_TT_delta = MO_TT - medTT)

if( T %in% is.na(df$MO_TT_delta)) {
 message(paste("WARNING, NA values found"))
 kable(df %>% filter(is.na(MO_TT_delta)))
 kable(df %>% filter(STATIONS_ID == 840))
}

```

## calculate the median of all deviations

```{r diff_to_ref_peryr}

devpyr = df %>% 
     group_by(yr,mo) %>% 
     summarize(med_MO_TT_diff = median(MO_TT_delta))
     
glimpse(devpyr %>% filter(yr == 2019))

```

# Let's plot!

```{r config_plots, fig.width = 10, include=FALSE}
library(ggplot2)

mytheme = theme_bw(base_size=20) + theme(
    ## ##text = element_text(family = "Decima WE", color = "grey20"),
    ## ## strip.background = element_blank(),
    ## ## strip.text = element_text(hjust = 0),
    ## ## panel.grid.major = element_line(colour="grey50",size=0.35),
    ## ## panel.grid.minor = element_blank(),
    ## ## plot.margin=unit(c(0,1,0,1),"cm"),
    ## legend.position="top",
    ## plot.caption=element_text(hjust=1,size=14,colour="grey30"),
    ## plot.subtitle=element_text(face="italic",size=14,colour="grey40"),
    axis.text.y = element_text(size=12),
    plot.title=element_text(size=18,face="bold")
## 
)

```


```{r plot_start_end, fig.width = 10}
library(ggplot2)

firstyr = min(devpyr$yr)
lastyr = max(devpyr$yr)

first = ggplot(devpyr %>% filter(yr == firstyr),aes(x=mo,y=med_MO_TT_diff))
first = first + geom_line(color="red",size=2)
first = first + geom_point(color="red",size=4)
first = first + xlab("Month")
first = first + ggtitle(paste(firstyr,": median difference of monthly avg temperature to (1961-1990)"))
first = first + ylab("Median(Temp(Monthly Avg) - Reference) / (Degree Celius)")
first = first + scale_x_continuous(breaks=seq(1,12,1))
first = first + mytheme

first

last = ggplot(devpyr %>% filter(yr == lastyr),aes(x=mo,y=med_MO_TT_diff))
last = last + geom_line(color="red",size=2)
last = last + geom_point(color="red",size=4)
last = last + xlab("Month")
last = last + ggtitle(paste(lastyr,"median difference of monthly avg temperature (1961-1990)"))
last = last + ylab("Median(Temp(Monthly Avg) - Reference) / (Degree Celius)")
last = last + scale_x_continuous(breaks=seq(1,12,1))
last
```

# Let's animate

Taking the years since 2000 first:

```{r anim_small, fig.width = 10, echo=FALSE}
library(gganimate)

finaldf = devpyr

plot = ggplot(finaldf,aes(x=mo,y=med_MO_TT_diff))
plot = plot + geom_line(color="red",size=2)
#plot = plot + geom_point(color="red",size=4)
plot = plot + xlab("Month")
plot = plot + ggtitle("{frame_time}: median difference of monthly avg temperature (1961-1990)")
plot = plot + ylab("Median(Temp-Diff(Monthly Avg)) [Degree Celius]")
plot = plot + scale_x_continuous(breaks=seq(1,12,1))
plot = plot + mytheme

anim = plot + transition_time(yr) + enter_reset() + enter_appear() + shadow_trail(color="gray",alpha = 0.5, size = 1)


mv = animate(anim, 
nframes=length(unique(finaldf$yr)),
               fps=5,
               renderer = gifski_renderer(),
               width=800,
               height=450,
               end_pause=50,
               rewind=FALSE)
mv
anim_save("saxony_full.gif",mv)
```

